\phantomsection
\addcontentsline{toc}{chapter}{\bf ACKNOWLEDGEMENTS}
\begin{center}
    \large{\bf ACKNOWLEDGEMENTS}
\end{center}

I would like to express my sincere gratitude to Dr. Chaklam Silpasuwanchai, my chairperson, for his guidance, encouragement, and valuable feedback throughout the development of this thesis. His advice has helped me refine the direction, structure, and technical foundation of this study.

I am also deeply thankful to my committee members, Dr. Chantri Polprasert and Dr. Attaphongse Taparugssanagorn, for their time, thoughtful suggestions, and constructive comments, which have greatly supported the improvement of this work.

I would also like to express my special thanks to Pranisaa Charnparttaravanit, a PhD student, for her continuous guidance and support throughout this research journey. She generously recommended relevant papers, helped me shape the problem statement, and brainstormed the methodology with me. Her insights and encouragement played an important role in helping me develop this thesis direction more clearly and confidently.

I would like to acknowledge the support of the AIT Brain Lab and the academic environment that has provided me with the opportunity to explore research at the intersection of artificial intelligence, human-computer interaction, and mental health support.

Finally, I am grateful to my family, friends, classmates, and everyone who encouraged me during this process. Their support, patience, and motivation have helped me continue working toward the completion of this thesis.