\setlength{\footskip}{8mm}

\chapter{EXPERIMENT}

This chapter presents the experimental design used to evaluate AIM-RGL in simulated MI-CBT counseling dialogue. The purpose of the experiment is to examine whether reward-guided lookahead search can improve therapeutic response quality when compared with baseline response generation methods. The chapter describes the experimental conditions, including the simulated client setup, compared systems, lookahead configurations, reward model backbones, and policy model backbones. It also explains the evaluation protocol and metrics used to assess therapeutic skill quality, MI-CBT fidelity, therapeutic alliance, change talk, sustain talk, and reward model performance. Since the implementation and evaluation are still in progress, the final section of this chapter is reserved for reporting the experimental results after they are obtained.

\section{Experimental Design and Conditions}

\subsection{Experiment Overview}

This experiment is designed to evaluate AIM-RGL in simulated MI-CBT counseling dialogue. The main purpose is to examine whether reward-guided lookahead search can improve therapeutic response generation when the simulated client shows resistance, ambivalence, or changing readiness for structured intervention. In this setting, a response is not evaluated only by whether it appears appropriate in the current turn, but also by whether it is expected to support a better near-future therapeutic trajectory.

The experiment compares AIM-RGL with selected baseline therapist systems under the same simulated client scenarios. Each system receives the same client profile, cognitive conceptualization information, difficulty setting, and dialogue context. The resulting therapist-client conversations are evaluated using counseling-specific metrics, including therapeutic skill quality, MI-CBT fidelity, therapeutic alliance, change talk, and sustain talk. The experiment also examines the effects of different lookahead horizons, reward model backbones, and policy model backbones.

The experiment is guided by the following research questions:

\begin{itemize}
    \item \textbf{RQ1:} Does reward-guided lookahead search improve therapeutic skill quality and MI-CBT fidelity compared with baseline systems?
    
    \item \textbf{RQ2:} Does reward-guided lookahead search strengthen therapeutic alliance compared with baseline systems?
    
    \item \textbf{RQ3:} Does reward-guided lookahead search promote earlier increases in change talk and reductions in sustain talk compared with baseline systems?
    
    \item \textbf{RQ4:} How do different lookahead horizons, including one-step, two-step, and three-step lookahead, affect reward-guided MI-CBT response quality?
    
    \item \textbf{RQ5:} How does the choice of policy model backbone affect the quality of MI-CBT therapeutic responses?
\end{itemize}

Based on these research questions, the experiment tests the following hypotheses:

\begin{itemize}
    \item \textbf{H1:} Reward-guided lookahead search will produce higher therapeutic skill quality and stronger MI-CBT fidelity than baseline methods.
    
    \item \textbf{H2:} Reward-guided lookahead search will lead to stronger therapeutic alliance compared with baseline methods.
    
    \item \textbf{H3:} Reward-guided lookahead search will promote earlier increases in change talk and earlier reductions in sustain talk compared with baseline methods.
    
    \item \textbf{H4:} Longer lookahead horizons will improve MI-CBT response quality by considering future client reactions and therapeutic trajectory.
    
    \item \textbf{H5:} The choice of policy model backbone will affect the quality of MI-CBT therapeutic responses in simulated resistant-client conversations.
\end{itemize}

This experiment is limited to simulated therapist-client dialogue. It does not involve real patients or human clinical participants, and it does not measure real clinical treatment outcomes.

\subsection{Experimental Conditions}

The experiment includes four main experimental factors: therapist system, lookahead horizon, reward model backbone, and policy model backbone. The therapist system comparison is used to evaluate whether AIM-RGL improves therapeutic response quality compared with selected baseline systems. The lookahead horizon comparison is used to examine whether considering more future client reactions improves response selection. The reward model backbone comparison is used to evaluate reward prediction performance, while the policy model backbone comparison is used to examine whether policy optimization results depend on the underlying response generation model.

The therapist systems compared in this experiment are CBT, CAMEL, FLASH, and AIM-RGL. The CBT baseline represents a structured cognitive-behavioral response generation approach. It focuses on CBT-style skills such as identifying thoughts, examining evidence, and supporting structured reflection or action. CAMEL is included as a baseline counseling agent for comparison \cite{lee2024cactus}. FLASH is included as a memory-guided MI-CBT baseline that uses previous client interactions to guide therapeutic strategy selection \cite{anonymous2026flash}. AIM-RGL is the proposed framework, which uses specialist therapist agents, client simulation, reward-guided lookahead search, reward model training, and policy optimization.

For AIM-RGL, the lookahead horizon is varied using one-step, two-step, and three-step lookahead settings. These are not treated as separate system names, but as different configurations of the same reward-guided lookahead mechanism. In one-step lookahead, each candidate therapist response is followed by one simulated client response before scoring. In two-step lookahead, the system expands the trajectory further to estimate a longer short-term therapeutic effect. In three-step lookahead, the system examines whether deeper trajectory evaluation provides additional improvement in response quality.

\begin{table}[htbp]
\centering
\caption{Experimental conditions.}
\begin{tabular}{L{4cm}L{9cm}}
\hline
\textbf{Condition Type} & \textbf{Experimental Settings} \\
\hline
Therapist system & CBT, CAMEL, FLASH, AIM-RGL \\
\hline
Lookahead horizon & One-step lookahead, two-step lookahead, three-step lookahead \\
\hline
Reward model backbone & DeBERTa-v3-large, ModernBERT-large, RoBERTa-large \\
\hline
Policy model backbone & LLaMA-3.1-8B-Instruct, Qwen-2.5-7B-Instruct-1M \\
\hline
\end{tabular}
\end{table}

The reward model backbone comparison includes DeBERTa-v3-large, ModernBERT-large, and RoBERTa-large. These models are used to examine how different encoder-based architectures affect scalar therapeutic reward prediction. The policy model backbone comparison includes LLaMA-3.1-8B-Instruct and Qwen-2.5-7B-Instruct-1M. These models are used to examine how the choice of response generation backbone affects optimized MI-CBT therapeutic response quality.

\subsection{Dataset and Client Simulation Setup}

The experiment uses simulated client scenarios based on the Patient-\(\Psi\)-CM dataset \cite{wang2024patientpsi}. The dataset contains CBT-based cognitive models and patient cases. Each case includes cognitive conceptualization information such as the client’s situation, automatic thoughts, emotions, behaviors, coping strategies, relevant history, intermediate beliefs, and core beliefs. These elements are used to ground the simulated client’s responses and maintain consistency across multi-turn dialogue.

The client simulator generates first-person client responses based on the current conversation history, the client’s cognitive conceptualization profile, and the client’s current stage of change. The stage of change is represented using three states: precontemplation, contemplation, and preparation. Precontemplation represents a resistant client who may minimize the problem, reject feedback, or show little interest in change. Contemplation represents an ambivalent client who recognizes both reasons for and against change. Preparation represents a client who is more open to structured reflection, planning, or CBT-style intervention.

Client difficulty is represented using easy, normal, and hard settings. Difficulty controls how frequently the client's openness level and stage of change are re-evaluated during the dialogue. Easy clients are expected to become more open more quickly, normal clients change more gradually, and hard clients remain resistant for longer. The openness judge periodically re-evaluates the simulated client's openness level and stage of change based on the recent conversation history. For easy clients, openness is re-evaluated every two turns. For normal clients, it is re-evaluated every four turns. For hard clients, it is re-evaluated every six turns. This interval-based design allows the simulation to represent different levels of resistance persistence and readiness development.

GPT-4o-mini is used for therapist-side response generation and client simulation. In AIM-RGL, the therapist-side generation module includes specialist agents for open-ended questioning, reflection, affirmation, summarization, and CBT intervention. These agents generate candidate responses before the reward-guided lookahead search stage. The reward model is trained to predict scalar therapeutic reward scores for candidate responses or trajectories. GPT-4o is used as a reward judge and evaluator for therapeutic quality dimensions, including therapeutic alliance, MI fidelity, and CBT competence. The same model, dataset, and client simulation setup are applied across compared systems to support controlled comparison.

\subsection{Reward Model and Policy Training Configuration}

This subsection describes the training configuration used for the reward model and the policy optimization stage. The reward model is trained before policy optimization because it provides the scalar therapeutic reward signal used to score candidate therapist responses during reward-guided lookahead search and GRPO. The methodological details of reward model training and GRPO are already described in Chapter 3; therefore, this subsection focuses only on the experimental configuration, dataset size, model backbones, and training setup.

For reward model training, 10 simulated client cases are selected from the Patient-\(\Psi\)-CM dataset. Each client case is evaluated under three difficulty modes: easy, normal, and hard. For each client and difficulty mode, 29 scored therapist turns are used. In the one-step lookahead setting, five candidate therapist responses are generated for each scored turn, corresponding to the main therapist agents. Therefore, the one-step reward dataset contains \(10 \times 3 \times 29 \times 5 = 4{,}350\) samples.

For deeper lookahead, the reward dataset is generated using beam-guided trajectory expansion. In the two-step lookahead setting, four judged trajectories are stored for each scored turn, producing \(10 \times 3 \times 29 \times 4 = 3{,}480\) samples. The three-step lookahead setting follows the same sample structure and also produces \(10 \times 3 \times 29 \times 4 = 3{,}480\) samples. In total, the planned reward model training dataset contains 11,310 samples across one-step, two-step, and three-step lookahead data.

The reward model is trained as a regression model. Its objective is to predict a scalar therapeutic reward score for each candidate response or lookahead trajectory. The reward targets are generated by the reward judge based on therapeutic dimensions such as therapeutic alliance, understanding and empathy, collaboration and structure, CBT skill and strategy, and MI quality and readiness. The dataset is divided into training, validation, and test sets using a 70\%, 15\%, and 15\% split. Mean squared error loss is used for reward model training because the reward target is represented as a continuous scalar score.

\begin{table}[h]
\centering
\caption{Reward model training configuration.}
\label{tab:reward-model-training-config}
\begin{tabular}{L{4.2cm}L{8.3cm}}
\hline
\textbf{Configuration} & \textbf{Setting} \\
\hline
Training objective & Predict scalar therapeutic reward score for candidate responses or trajectories \\ \hline
Training client cases & 10 simulated client cases \\ \hline
Difficulty modes & Easy, normal, hard \\ \hline
Scored turns per client and mode & 29 therapist turns \\ \hline
Data split & Train 70\%, validation 15\%, test 15\% \\ \hline
Loss function & Mean squared error loss \\ \hline
Reward model backbones & DeBERTa-v3-large, ModernBERT-large, RoBERTa-large \\ \hline
One-step dataset \(D_1\) & \(10 \times 3 \times 29 \times 5 = 4{,}350\) samples \\ \hline
Two-step dataset \(D_2\) & \(10 \times 3 \times 29 \times 4 = 3{,}480\) samples \\ \hline
Three-step dataset \(D_3\) & \(10 \times 3 \times 29 \times 4 = 3{,}480\) samples \\ \hline
Total reward dataset & 11,310 samples \\
\hline
\end{tabular}
\end{table}

After the reward model is trained, it is used to support policy optimization through GRPO. In this stage, the policy model generates a group of candidate therapist responses for the same conversation context. The trained reward model scores each candidate response, and the group-level reward distribution is used to compute relative advantages. The policy model is then optimized to increase the likelihood of higher-reward therapeutic responses while remaining close to the reference model through a KL penalty.

The GRPO stage compares two policy model backbones: LLaMA-3.1-8B-Instruct and Qwen-2.5-7B-Instruct-1M. Low-rank adaptation is used to reduce the cost of policy optimization. The planned LoRA configuration uses a rank of 8 and a LoRA alpha of 32. A fixed copy of the initial policy backbone is used as the reference model for calculating the KL divergence penalty. The number of generated responses per context is treated as the GRPO group size. The exact learning rate, batch size, number of epochs, and number of GRPO training client cases will be finalized during implementation based on available computational resources.

\begin{table}[h]
\centering
\caption{GRPO policy optimization configuration.}
\label{tab:grpo-training-config}
\begin{tabular}{p{4.2cm}p{8.3cm}}
\hline
Configuration & Setting \\
\hline
Policy optimization method & Group Relative Policy Optimization \\
Policy model backbones & LLaMA-3.1-8B-Instruct, Qwen-2.5-7B-Instruct-1M \\
Reward source & Trained reward model \\
Training signal & Group-normalized reward advantage \\
Number of generations & Treated as the GRPO group size \\
LoRA rank & 8 \\
LoRA alpha & 32 \\
Reference model & Fixed copy of the initial policy backbone \\
Regularization & KL divergence penalty against the reference model \\
GRPO training client cases & To be finalized during implementation \\
Learning rate, batch size, and epochs & To be finalized during implementation \\
\hline
\end{tabular}
\end{table}

The reward model and policy optimization stages are separated to make the experimental comparison clearer. The reward model training stage evaluates whether the system can learn to approximate therapeutic reward scores from lookahead data. The GRPO stage then evaluates whether this learned reward signal can improve the therapeutic response generation policy across different policy model backbones.

\section{Experimental Results}

This section presents the current experimental results and describes how the final results will be reported after the full AIM-RGL implementation and evaluation are completed. At the current stage, the available results are preliminary baseline results comparing CBT and FLASH under easy, normal, and hard simulated client settings. The final version of this section will include the complete comparison between CBT, CAMEL, FLASH, and AIM-RGL, together with lookahead horizon, reward model backbone, and policy model backbone comparisons.

\subsection{Therapeutic Response Quality Results}

The preliminary evaluation compares CBT and FLASH using therapist skill dimensions and broader therapeutic quality metrics. The therapist skill dimensions include approach, discovery, focus, strategy, understanding, and interpersonal collaboration. Table~\ref{tab:prelim-therapist-skills} summarizes the preliminary therapist skill results across easy, medium, and hard simulated client levels.

\begin{table}[h]
\centering
\caption{Preliminary therapist skill results for CBT and FLASH.}
\label{tab:prelim-therapist-skills}
\footnotesize
\setlength{\tabcolsep}{3.5pt}
\begin{tabular}{llcccccc}
\hline
Level & System & Approach & Discovery & Focus & Strategy & Understanding & Collaboration \\
\hline
Easy & CBT & 4.00 & 4.00 & 3.60 & 4.00 & 4.80 & 4.00 \\
Easy & FLASH & 4.00 & 3.80 & 3.60 & 4.00 & 4.60 & 3.60 \\
\hline
Medium & CBT & 4.00 & 4.00 & 3.20 & 4.00 & 4.40 & 4.00 \\
Medium & FLASH & 4.00 & 3.40 & 3.40 & 4.00 & 4.80 & 3.40 \\
\hline
Hard & CBT & 4.00 & 3.80 & 3.40 & 4.00 & 4.20 & 3.20 \\
Hard & FLASH & 4.00 & 3.40 & 3.00 & 3.80 & 5.00 & 3.30 \\
\hline
\end{tabular}
\end{table}

The preliminary results show that CBT obtains stronger scores than FLASH on several therapist skill dimensions, especially discovery and collaboration. CBT also remains relatively stable across difficulty levels. FLASH shows stronger understanding scores in medium and hard settings, but its discovery, focus, and collaboration scores are lower in several conditions. This suggests that FLASH may generate responses that appear understanding, but may provide less structured therapeutic progression in more difficult simulated client conditions.

Table~\ref{tab:prelim-quality-alliance} summarizes the preliminary therapeutic quality and alliance results. The evaluation includes CTRS, WAI, MITI, and alliance scores across easy, normal, and hard simulated client settings.

\begin{table}[h]
\centering
\caption{Preliminary therapeutic skills and alliance results.}
\label{tab:prelim-quality-alliance}
\small
\setlength{\tabcolsep}{8pt}
\begin{tabular}{llcccc}
\hline
Level & System & CTRS & WAI & MITI & Alliance \\
\hline
Easy & CBT & 3.73 & 3.67 & 4.22 & 3.84 \\
Easy & FLASH & 3.47 & 3.43 & 4.24 & 3.64 \\
\hline
Normal & CBT & 3.65 & 3.55 & 4.04 & 3.79 \\
Normal & FLASH & 2.77 & 3.23 & 3.92 & 3.27 \\
\hline
Hard & CBT & 3.61 & 3.71 & 3.88 & 3.77 \\
Hard & FLASH & 2.82 & 3.06 & 3.80 & 3.24 \\
\hline
\end{tabular}
\end{table}

The preliminary results indicate that CBT performs better than FLASH on CTRS, WAI, and alliance across all difficulty levels. FLASH obtains a slightly higher MITI score in the easy condition, but CBT performs better on MITI in the normal and hard conditions. This suggests that CBT provides stronger structured therapeutic competence and alliance in the current preliminary evaluation, while FLASH may preserve some MI-related qualities in easier settings. However, these findings should be interpreted only as preliminary baseline results because the full AIM-RGL comparison has not yet been completed.

\subsection{Ablation and Model Comparison Results}

The ablation and model comparison results will be added after the full AIM-RGL experiment is completed. This part of the experiment will examine whether reward-guided lookahead search improves therapeutic response quality compared with the baseline systems. It will also analyze how different lookahead horizons affect response quality.

The lookahead comparison will evaluate one-step, two-step, and three-step lookahead settings. The purpose of this comparison is to determine whether deeper lookahead improves therapeutic response selection by considering more future client reactions, or whether the benefit becomes limited after a certain depth. The final results will be reported using the same therapeutic evaluation metrics, including CTRS, WAI, MITI, alliance, change talk, and sustain talk.

\begin{table}[h]
\centering
\caption{Planned lookahead horizon comparison.}
\label{tab:lookahead-comparison-placeholder}
\begin{tabular}{lccccc}
\hline
Lookahead Horizon & CTRS & WAI & MITI & Change Talk & Sustain Talk \\
\hline
One-step lookahead & -- & -- & -- & -- & -- \\
Two-step lookahead & -- & -- & -- & -- & -- \\
Three-step lookahead & -- & -- & -- & -- & -- \\
\hline
\end{tabular}
\end{table}

The reward model backbone comparison will evaluate DeBERTa-v3-large, ModernBERT-large, and RoBERTa-large. Since the reward model predicts a scalar therapeutic reward score, the comparison will use regression and ranking-based metrics such as mean squared error, mean absolute error, Pearson correlation, and Spearman rank correlation.

\begin{table}[h]
\centering
\caption{Planned reward model backbone comparison.}
\label{tab:reward-backbone-placeholder}
\begin{tabular}{lcccc}
\hline
Reward Model Backbone & MSE & MAE & Pearson $r$ & Spearman $\rho$ \\
\hline
DeBERTa-v3-large & -- & -- & -- & -- \\
ModernBERT-large & -- & -- & -- & -- \\
RoBERTa-large & -- & -- & -- & -- \\
\hline
\end{tabular}
\end{table}

The policy model backbone comparison will evaluate LLaMA-3.1-8B-Instruct and Qwen-2.5-7B-Instruct-1M after policy optimization using the trained reward model. This comparison will show whether the quality of MI-CBT response generation depends on the underlying policy model backbone.

\begin{table}[h]
\centering
\caption{Planned policy model backbone comparison.}
\label{tab:policy-backbone-placeholder}
\begin{tabular}{lccccc}
\hline
Policy Backbone & CTRS & WAI & MITI & Change Talk & Sustain Talk \\
\hline
LLaMA-3.1-8B-Instruct & -- & -- & -- & -- & -- \\
Qwen-2.5-7B-Instruct-1M & -- & -- & -- & -- & -- \\
\hline
\end{tabular}
\end{table}

\subsection{Summary of Findings}

The current preliminary results provide an initial baseline comparison between CBT and FLASH. CBT shows stronger overall performance in structured therapist skill quality, CTRS, WAI, and alliance across easy, normal, and hard simulated client settings. FLASH shows competitive MI-related performance in the easy setting, but its scores decrease more clearly in normal and hard simulated client conditions.

These preliminary findings support the need for a response selection mechanism that can balance structured CBT competence with MI-consistent readiness handling. The final AIM-RGL evaluation will examine whether reward-guided lookahead search can improve this balance by comparing candidate MI-CBT responses according to their expected near-future therapeutic effects. The final summary will be updated after AIM-RGL, lookahead horizon, reward model backbone, and policy backbone results are completed.
